A primer to numerical simulations: The perihelion motion of Mercury

### 2018-01-12 tex/MercuryNumerical.tex
My revision of the paper

\author{
	C.~K\"orber$^{1}$,
	I.~Hammer$^{1}$,
	J.-L.~Wynen$^{1}$,
	J.~Heuer$^{2}$,
	C.~M\"uller$^{3}$ and
	C.~Hanhart$^{1}$
}
\address{
	$^1$ \textit{Institute f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^2$ \textit{Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^3$ \textit{JuLab, Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}
}
\ead{c.koerber@fz-juelich.de, c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science.
In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization of the results.
In addition a method is discussed how to estimate the size of the effect a priori that at the same time also can be employed to
double check the results found numerically.
\end{abstract}

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise, e.g., since there are too many particle participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \texttt{VPython}.
\end{itemize}
The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check, if the results of the simulation are sensible.

We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Schülerakademie Teilchenphysik", where this course was tested successfully on two groups consisting of in total 24
German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Trajectories, velocities, accelerations and Newtons second law}\label{sec:tva}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A physical system is understood, if the assumed force leads to the trajectories observed.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space) the location is parametrized by a single three vector $\vec r(t)$.
As will become clear below, in order to describe and control the dynamics of a physical object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced with the last expression a common short hand notation for time derivatives.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law.
Said differently: a force $\vec F$ shows up by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation}
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as the rocket rises.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation}
Note that in general the force could depend also on the time or the velocity.
We here restrict ourselves to the case relevant for our example, where the force depends on the location only.
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general a differential equation of $n^{\rm th}$ degree (where the highest derivative is $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
}.

For this procedure to work $\Delta t$ must be sufficiently small.
What this means depends on the process studied.
One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\ .
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newtons law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec r(t)|$ denotes the distance between sun and Mercury, when we assume that the sun is infinitely more heavy than the planet and located at the center of the coordinate system.
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\label{rsdef}
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
With this Newtons second law reads
\begin{equation}
	\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, .
\label{eq:newton}
\end{equation}
In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the sun, since a given force can capture only bodies with not too large momenta.
The students may study those scenarios within their simulations  by varying the start velocity while keeping the start location fixed.

The bounded orbits that emerge from a potential that scales as $1/r$ are elliptic and fixed in space.
In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
However, when a potential that vanishes for $r\to \infty$ deviates from $1/r$, the perihelion moves.
Because of this, the behavior of the perihelion is a very sensitive probe of the gravitational potential.

The observed perihelion motion of Mercury is nowadays determined as
$$(574.10\pm 0.65)'' \ \mbox{per
100 earth years}$$
where the symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}\delta \Theta_M = (43)'' \ \mbox{in 100 earth years}\label{delT}\end{equation}
 remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that appears in Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
However, besides $r_S$, which is the characteristic length scale for the gravitational potential
of the sun, one can define one more parameter, which is characteristic for the dynamics of Mercury, namely
\begin{equation}
a^2=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
\label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $a^2$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
for the modified potential
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand $a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we
already indicate that below we start the simulation at the perihelion) and the closest distance between sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be safely neglected.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
However, since we have with the perihelion shift (see Eq.~(\ref{delT})) only a single number to study,
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
At the same time the parameters can also be calculated from the assumed underlying theory and the actual values
are in general specific for this theory. It turns out that
General relativity gives~\cite{Einstein}
\begin{equation}\label{eq:art-prediction}
	\alpha = 0 \, , \quad \beta = 3 \, .
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

%For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
%Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
%\begin{equation}
%	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
%\end{equation}
%where $T_V$ ($T_M$) denotes the length of a venus (Mercury) year. This precisely describes the experimental value of $9''$.

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{} \cite{Python} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
In  \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

To start the simulation one needs the "initial" distance and velocity [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (Mercury) in a two-dimensional plane.

Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we
use as initial (for $t=0$) parameters the values at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$~km and $ \abs{\vec v_M(0)} = 59$~km/s\cite{MercuryFactSheet}.
Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one Mercury year is given by $T_M=88\,T_0$.
With this choice, the initial distance of Mercury to the sun, the size of the initial velocity of Mercury and the acceleration pre-factors become
\begin{eqnarray}
r_{MS}(0)   &=& 4.6 R_0 \, , \quad
v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
a_M(r_{MS}) &=& 1.01 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
\, .
\end{eqnarray}

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}%
		Sun Mercury system with relevant vectors.
		Because Mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%

>>>>

\author{
	C.~K\"orber$^{1}$,
	I.~Hammer$^{1}$,
	J.-L.~Wynen$^{1}$,
	J.~Heuer$^{2}$,
	C.~M\"uller$^{3}$ and
	C.~Hanhart$^{1}$
}
\address{
	$^1$ \textit{Institut f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^2$ \textit{Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
	$^3$ \textit{JuLab, Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}
}
\ead{c.koerber@fz-juelich.de, c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science.
In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization of the results.
In addition a method is discussed how to estimate the size of the effect a priori. Which allows to double check the results found numerically.
\end{abstract}

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
The course \todo{what course?} as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check whether the results of the simulation are sensible.

We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep knowledge of Differential Geometry. For the study outlined here, however, very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017\todo{do foreigners know what that means? what are typical ages of students? did grade 13 even exist in 2015?}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Trajectories, velocities, accelerations and Newton's second law}\label{sec:tva}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A physical system is said to be understood if the assumed force acting on it leads to the observed trajectories.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $a(t)$ in order to describe and control its dynamics; this will become clear below. The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + \ldots \ . \label{eq:vdef}
\end{equation}
Similarly, the acceleration describes a change in velocity, i.e.
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + \ldots\ . \label{eq:adef}
\end{eqnarray}
The dots in these expressions indicate additional terms that may be expressed in terms of higher powers in $\Delta t$. While they are needed in general, for sufficiently small $\Delta t$ they can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced a common short hand notation for time derivatives in the last expression.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced the second derivative in the second line.

It was Newton who observed that if a body is at rest, it will remain at rest, and if it is in motion it will remain in motion at a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newton's first law.
Formulated differently: a force $\vec F$ expresses itself by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation}
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as fuel is burned.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation}
Note that in general the force could depend also on the time or the velocity.
Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use the force to calculate $\vec a(t) = \vec F(\vec r(t)) / m$, see Eq.~(\ref{eq:newton2}).
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$.
\item Use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some starting time, but one may as well pick two locations at different times.%
}.

This procedure can only work if $\Delta t$ is sufficiently small.
What this means depends on the process one studies.
One way to estimate whether $\Delta t$ is small enough, is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that we neglected reads $(1/2)a(t){(\Delta t)}^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

The bounded orbits that emerge from a potential that scales as $1/r$ (which corresponds to a force that scales as $1/r^2$) are elliptic and fixed in space.
In particular the point of closest approach of the planet to the Sun, the perihelion, does not move.
However, when a potential that vanishes for $r\to \infty$ deviates from $1/r$, the perihelion does move.
This makes the behavior of the perihelion a very sensitive probe of the gravitational potential.

The observed perihelion motion of Mercury is nowadays determined as \todo{reference?}
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (43)'' \ \mbox{in 100 earth years} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from a $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_a^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(r_a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size. Furthermore, any term of higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

% For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
% Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
% \begin{equation}
% 	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
% \end{equation}
% where $T_V$ ($T_M$) denotes the length of a venus (Mercury) year. This precisely describes the experimental value of $9''$.

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular the perihelion motion of Mercury.
We choose \python{}~\cite{Python} as programming language because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
In~\ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

To start the simulation one needs the "initial" distance and velocity [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
As described above, here one can simplify the problem by placing one object (the "infinitely" heavy Sun) in the center of the coordinate system and keep it fixed.

Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we
use the values at the perihelion with $\abs{\vec r_{MS}(0)} = 46 \cdot 10^6\,\mathrm{km}$ and $\abs{\vec v_M(0)} = 59\,\mathrm{km/s}$\cite{MercuryFactSheet} as initial (for $t=0$) parameters.
The computer does not understand physical units, hence one has to express each variable in an appropriate unit.
Both for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
by expressing distances in $R_0 = 10^{10}\,\mathrm{m}$ and time intervals in days, $T_0 = 1\,\mathrm{d}$, where d refers to earth days. One Mercury year is given by $T_M=88\,T_0$.
With this choice, the initial distance of Mercury to the Sun, the size of the initial velocity of Mercury and the acceleration prefactor (see Eq.~(\ref{eq:newton_art})) become
\begin{eqnarray}
r_{MS}(0)   &=& 4.6 R_0 \, , \quad
v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r^2} = 1.01 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
\end{eqnarray}

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}%
		Sun-Mercury system with relevant vectors.
		Mercury is at its perihelion, therefore its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%
